Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-02-13 curvature.tex
flesh out second half of lem:degree_max_delta_adjacent, s/node/vertex; s/nodes/vertices

For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
A \emph{parent (sub)tree} of a subtree $U$ is the smallest subtree strictly containing $U$.
A \emph{parent edge} of a subtree $U$ is the edge connecting $U$ to the rest of the tree.
The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
A \emph{ladder tree} is a tree such that every internal node has a leaf as a direct descendant.
LCA of subtrees.

@@

In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of nodes in the graph.
Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.

@@

\begin{lemma}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $T$ can be obtained from $S$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$. \todo{EM: if we move R to an ancestor (e.g. the root) then the subtree $V$ in $T$ contains both $R$ and $U$. $R$ and $U$ are disjoint in $T$, though so I dropped one excluding.}
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{lemma}
\begin{proof}
The set of permissible SPR moves changes in four different ways due to the movement of $R$:
1) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
2) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,\todo{let's choose "node" or "vertex"}
3) $R$'s parent subtree may now be moved into $U$, and
4) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original SPR operation on $R$.

@@

\begin{corollary}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ nodes in the SPR graph be $q$ and $p$ respectively.
Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
\end{corollary}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}

@@

	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf \todo{Hm, I'm not sure where this additional leaf is! EM: does this make more sense?} to balance the product at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	If we move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
Similarly, if we do not move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$

The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
This is is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximum. \todo{This sort of thing does not appear true in general. Please explain why it works here. EM: Is this better? Think of $\frac{x}{1+x}$ which is an increasing function.}
	That is, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = \degree{T} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.

>>>>

For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
A \emph{parent (sub)tree} of a subtree $U$ is the smallest subtree strictly containing $U$.
A \emph{parent edge} of a subtree $U$ is the edge connecting $U$ to the rest of the tree.
The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
A \emph{ladder tree} is a tree such that every internal vertex has a leaf as a direct descendant.
LCA of subtrees.

@@

In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.

@@

\begin{lemma}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $T$ can be obtained from $S$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate vertices on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate vertices on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{lemma}
\begin{proof}
The set of permissible SPR moves changes in four different ways due to the movement of $R$:
1) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent vertex,
2) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent vertex,
3) $R$'s parent subtree may now be moved into $U$, and
4) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original SPR operation on $R$.

@@

\begin{corollary}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ vertices in the SPR graph be $q$ and $p$ respectively.
Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
\end{corollary}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target vertices.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}

@@

	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	If we move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
Similarly, if we do not move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
Therefore claim (1) holds.

The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
By (1), we have that $\degree{S} - \degree{T} \le \frac{1}{2} (n-2)^2 $.
Thus, $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
This bound is maximized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.
